---
title: "Eviewで学ぶ応用ファイナンス"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(grid)
library(jpeg)
library(knitr)
library(xtable)
library(bizdays)
library(zoo)
library(stats)
library(moments)
library(tseries)
library(RQuantLib)



# exercise enviorment setting
create.calendar("business_day", weekdays=c("saturday", "sunday"))
series_date=bizseq("2010-01-01", "2015-12-31", "business_day")

x=c(2000,2020) #initial value
set.seed(100) #random seed
random_3=rnorm(length(series_date))
for(i in 1:length(series_date)){
  x=append(x,0.3*tail(x,2)[2]+0.6*tail(x,2)[1]+4*random_3[i])
}
x_ts=zoo(round(x,2),order.by = series_date)

descriptive=function(x){
  jb=tseries::jarque.bera.test(x)
  results=data.frame(
  'Mean'=mean(x),
  'Median'=median(x),
  'Maximum'=max(x),
  'Minimum'=min(x),
  'Std.Dev'=sd(x),
  'Skewness'=moments::skewness(x),
  'Kurtosis'=moments::kurtosis(x),
  'Jarque-Bera'=jb$statistic,
  'Probability'=jb$p.value)
  rownames(results)=NULL
  return(results)
}

dcam=5000
set.seed(1111)
random_2=rnorm(length(series_date))
for(i in 1:length(series_date)){
  dcam=append(dcam,tail(dcam,1)+6*random_2[i])
}
dcam_ts=zoo(round(dcam,2),order.by = series_date)

rd=na.omit(diff(log(dcam_ts)))

img <- readJPEG("IMG_3652.JPG")

knitr::opts_chunk$set(echo = FALSE)

```
## Chapter One: Statistical Analysis

- Author: Xie HE
- Graduate School of Economics, Kobe University
- Hamori Seminar



```{r,echo=FALSE}
grid.raster(img)
```



## Preface{data-progressive=FALSE}  

### Introduction  

This tutorial intoduces how to use R programming to implement analysis by Eviews in chapter 1 statistical analysis of book Eviewで学ぶ応用ファイナンス. In chapter one, there are four sections.

- 1.1 the fundamental knowledge of time series 

- 1.2 model estimation by OLS

- 1.3 estimation of AR model  

- 1.4 introduction of Kalman filter (see next week)


The eviews codes in book have been rewritten in R.

### Necessary R packages  
```{r,message=FALSE,echo=TRUE,eval=FALSE}
library(bizdays)
library(zoo)
library(moments)
library(tseries)
library(RQuantLib)
```


### For more details about above packages
- bizdays  
<https://cran.r-project.org/web/packages/bizdays/bizdays.pdf>  
- zoo  
<https://cran.r-project.org/web/packages/zoo/zoo.pdf>  
- moments  
<https://cran.r-project.org/web/packages/moments/moments.pdf>  
- tseries   
<https://cran.r-project.org/web/packages/tseries/tseries.pdf>  
- RQuantLib  
<https://cran.r-project.org/web/packages/RQuantLib/RQuantLib.pdf>

## 1.1 The fundamental knowledge of time series   

### Generating time series

We not use real stock prices here but instead simulate two virtual stock prices following unit root process for analysis.  
- Stock market index  
- Stock prices of DCAM  
Period: 2010-01-01~2015-12-31 (only business day)

### Generating a date series
Firstly, we need to generate the data series from January 1st 2010 to December 31th 2015, and since only business days in the period are wanted, we use the `create.calendar()` function from `bizdays` package.   

```{r , echo = TRUE}
create.calendar("business_day", weekdays=c("saturday", "sunday"))
series_date=bizseq("2010-01-01", "2015-12-31", "business_day")
length(series_date)
```

What should we do if not only don't want weekdays but also don't want holidays?
then you need `load_quantlib_calendars()` in `bizdays` package so that you can load a calender of one country from `RQuantLib` package, and run following codes:

```{r  , echo = TRUE}
# load calendar of Japan from RQuantLib
load_quantlib_calendars(ql_calendars = 'Japan', from="2010-01-01",
                        to="2015-12-31")
# set the calendar to bizseq function
series_date_jp=bizseq("2010-01-01", "2015-12-31", "QuantLib/Japan")
length(series_date_jp)
```

### Exercise 
Please generate a date series without weekdays and holidays in China from 2010-01-01 to 2015-12-31, and check the length of series.

```{r a, exercise=TRUE, exercise.eval=TRUE,message=FALSE}

```

```{r a-solution, exercise.reveal_solution = FALSE}
load_quantlib_calendars(ql_calendars = 'China', from="2010-01-01",
                        to="2015-12-31")
series_date_cn=bizseq("2010-01-01", "2015-12-31", "QuantLib/China")
length(series_date_cn)
```

```{r a-hint-1}
ql_calendars='China'
```

```{r a-hint-2}
#Here is solution.
load_quantlib_calendars(ql_calendars = 'China', from="2010-01-01",
                        to="2015-12-31")
series_date_cn=bizseq("2010-01-01", "2015-12-31", "QuantLib/China")
length(series_date_cn)
```

### Generating stock market index
Next, we are going to simulate a time series following unit root process as stock market index.
$$
\begin{equation}\nonumber
    tx_t=tx_{t-1}+u_{2t}, u_{2t} \sim N(0,9)
    \end{equation}
$$ 
In here, you may need random gerneration, try `rnorm()` function:
```{r, echo = TRUE}
rnorm(5,mean=0,sd=1) # random generation
```
Then you can get 5 random numbers from standard normal distribution.

However, there is one thing you need to notice is that if you don't set the random seed using `set.seed()`, you will get different random numbers each time you run the generating process.

Have a try (just run the code):
```{r b, exercise=TRUE, exercise.eval=TRUE}
rnorm(5,mean=0,sd=1)
```

###  
Here are some other functions about normal distribution: 
```{r, echo = TRUE}
dnorm(0,mean=0,sd=1) # probability density function
pnorm(0,mean=0,sd=1) # cumulative distribution function
qnorm(0.5,mean=0,sd=1) # quantile function
```
Then, using `rnorm`, we can generate the stock market index:
```{r, echo = TRUE}
tx=1000
set.seed(123456)
random=rnorm(length(series_date))
for(i in 1:length(series_date)){
  tx=append(tx,tail(tx,1)+3*random[i])
  }
tx_ts=zoo(round(tx,2),order.by = series_date)
plot(tx_ts)
```

### Generating stock prices of DCAM
And you can also simulate another time series as stock prices of DCAM based on the same method.
$$
\begin{equation}\nonumber
    dcam_t=dcam_{t-1}+u_{1t}, u_{1t} \sim N(0,36)
    \end{equation}
$$ 
```{r, echo = TRUE}
dcam=5000
set.seed(1111)
random_2=rnorm(length(series_date))
for(i in 1:length(series_date)){
  dcam=append(dcam,tail(dcam,1)+6*random_2[i])
}
dcam_ts=zoo(round(dcam,2),order.by = series_date)
plot(dcam_ts)
```

### Exercise 
Please generate a time series from 2010-01-01 to 2015-12-31 (without weekdays) which following:
$$
\begin{equation}\nonumber
    x_t=0.3x_{t-1}+0.6x_{t-2}+u_{t}, u_{t} \sim N(0,16)
    \end{equation}
$$ 
Notice:  
- The two initial values are given as 2000 and 2020.    
- Set random seed as 100.  
- The date series (`series_date`) has already been defined, just use it directly.   
- Define the variable name of time series (zoo object) as 'x_ts'
```{r c, exercise=TRUE, exercise.eval=TRUE}

```

```{r c-hint-1}
# x_t-1
tail(x,2)[2]

# x_t-2
tail(x,2)[1]
```

```{r c-hint-2}
#Here is solution.
x=c(2000,2020) #initial value
set.seed(100) #random seed
random_3=rnorm(length(series_date))
for(i in 1:length(series_date)){
  x=append(x,0.3*tail(x,2)[2]+0.6*tail(x,2)[1]+4*random_3[i])
}
x_ts=zoo(round(x,2),order.by = series_date)
plot(x_ts)
```

```{r c-solution, exercise.reveal_solution = FALSE}
x=c(2000,2020) #initial value
set.seed(100) #random seed
random_3=rnorm(length(series_date))
for(i in 1:length(series_date)){
  x=append(x,0.3*tail(x,2)[2]+0.6*tail(x,2)[1]+4*random_3[i])
}
x_ts=zoo(round(x,2),order.by = series_date)
plot(x_ts)
```

Please check if your plot just looks like:  
```{r,echo=FALSE}
x=c(2000,2020) #initial value
set.seed(100) #random seed
random_3=rnorm(length(series_date))
for(i in 1:length(series_date)){
  x=append(x,0.3*tail(x,2)[2]+0.6*tail(x,2)[1]+4*random_3[i])
}
x_ts=zoo(round(x,2),order.by = series_date)
plot(x_ts)
```


### Calculating return series
Here are return series of two series that are calculated by first difference in log.
```{r, echo = TRUE}
rd=na.omit(diff(log(dcam_ts)))
rt=na.omit(diff(log(tx_ts)))
par(mfrow=c(2,1))
plot(rt)
plot(rd)
```

### Descriptive statistics
Next, we can check the descriptive statistics for series by using function `descriptive()` defined by ouselves. Here is the function: 
```{r echo=TRUE}
descriptive=function(x){
  jb=tseries::jarque.bera.test(x)
  results=data.frame(
  'Mean'=mean(x),
  'Median'=median(x),
  'Maximum'=max(x),
  'Minimum'=min(x),
  'Std.Dev'=sd(x),
  'Skewness'=moments::skewness(x),
  'Kurtosis'=moments::kurtosis(x),
  'Jarque-Bera'=jb$statistic,
  'Probability'=jb$p.value)
  rownames(results)=NULL
  return(results)
}
```

Here are results of DCAM stock prices.
```{r,echo=TRUE}
kable(descriptive(dcam_ts))
```
Here are results of DCAM stock returns.
```{r,echo=TRUE}
kable(descriptive(rd))
```

Check the results of `x_ts` defined before.
```{r d, exercise=TRUE, exercise.eval=TRUE}

```

### Histogram of DCAM stock prices

Next, we are going to show the histogram of two series.  
Here is the histogram of DCAM stock prices.
```{r, echo = TRUE}
hist(dcam_ts,breaks = seq(4500,5500,length.out=50),freq = T)
```

Here is the histogram of DCAM stock returns.
```{r, echo = TRUE}
hist(rd,breaks = seq(-0.004,0.004,length.out=50),freq = T)
```

Check the results of `x_ts`.
```{r e, exercise=TRUE, exercise.eval=TRUE}

```

### Normal distribution and t distribution

### Testing a mean using t-test
In this section, we used two-tail t-test for if mean equal 0:
$$
\begin{equation}\nonumber
    H_0:\mu=0,H_1:\mu \neq 0
    \end{equation}
$$

t-statistic:
$$
\begin{equation}\nonumber
    t=\frac{\overline{X}-0}{\frac{s}{\sqrt{n}}}
    \end{equation}
$$

Conducting t-test in R by using `t.test()` function:
```{r, echo=TRUE}
t.test(rd,mu=0) # alternative = c("two.sided", "less", "greater")
```

t statics and p-value can be calculated by ourselves.
```{r  echo=TRUE}
mu=mean(rd)
sd=sd(rd)
obs=length(rd)
h0=0
t=(mu-h0)/(sd/sqrt(obs))
p=pt(t,(obs-1))*2 #p=(1-pt(t,(obs-1)))*2, if t>0
c('t-value'=t,'p-value'=p)
```

#### Exercise 
Conducting a one-tail t-test for mean of `rd`, calculate t-statistic and p-value by yourself and vertify your answer by using `t.test()`:
$$
\begin{equation}\nonumber
    H_0:\mu>0,H_1:\mu\leq 0
    \end{equation}
$$

```{r f, exercise=TRUE, exercise.eval=TRUE}
mu=mean(rd)
sd=sd(rd)
obs=length(rd)
h0=0
```

```{r f-hint-1}
# only p-value changed
```

```{r f-hint-2}
#Here is solution.
mu=mean(rd)
sd=sd(rd)
obs=length(rd)
h0=0
t=(mu-h0)/(sd/sqrt(obs))
p=pt(t,(obs-1))
c('t-value'=t,'p-value'=p)
t.test(rd,mu=0,alternative='less') 
```

```{r f-solution, exercise.reveal_solution = FALSE}
mu=mean(rd)
sd=sd(rd)
obs=length(rd)
h0=0
t=(mu-h0)/(sd/sqrt(obs))
p=pt(t,(obs-1))
c('t-value'=t,'p-value'=p)
t.test(rd,mu=0,alternative='less') 
```

### Illustration of central limit theorem using Monte-Carlo Simulation

### Central limit theorem
The central limit theorem states that if you have a population with mean μ and standard deviation σ and take sufficiently large random samples from the population with replacementtext annotation indicator, then the distribution of the sample means will be approximately normally distributed.  

- Central limit theorem  
If $X_1, X_2, \dots, X_n$ are random samples and $i.i.d \quad (\mu, \sigma^2)$, then:  
$Y_n=\frac{\sqrt{\hat{X}-\mu}}{\sigma}$,where $\overline{X}_n=\frac{\sum^{n}_{i=1}X_i}{n}$,   
$s.t. \quad Y_n \overset{D}{\rightarrow} N(0,1)$

### Monte-Carlo Simulation
Monte Carlo methods, or Monte Carlo experiments, are a broad class of computational algorithms that rely on repeated random sampling to obtain numerical results.

```{r echo=TRUE,eval=FALSE}
obs=10
mu=170
sd=5
reps=1000 
hgt=c()
m=c()
se=c()
t=c()
set.seed(1234)
random_3=matrix(rnorm(reps*obs),reps,obs)
for(i in 1:reps){
  for(j in 1:obs){
    hgt[j]=random_3[i,j]*sd+mu
  }
  m[i]=mean(hgt)
  se[i]=sd(hgt)/sqrt(obs)  
  t[i]=(mu-m[i])/se[i]
}
hist(t,breaks = seq(-7,7,0.2),freq = F)
xfit=seq(-7,7,0.2)
yfit_1<-dnorm(xfit,mean=mean(t),sd=sd(t)) 
lines(xfit, yfit_1, col="blue", lwd=2) 
yfit_2<-dt(xfit,obs-1) 
lines(xfit, yfit_2, col="red", lwd=2) 
legend(x = "topright", legend=c("normal distribution", "t distribution"),
       col=c("blue", "red"),lty=c(1,1),cex=1)
```


```{r, echo=FALSE}
sliderInput("obs", "Number of samples:", min = 10, max = 1000, value = 50)
sliderInput("reps", "Sample size:", min = 100, max = 2000, value = 100)
plotOutput("distPlot")
```

```{r, context="server"}
output$distPlot <- renderPlot({
obs=input$obs
mu=170
sd=5
reps=input$reps 
hgt=c()
m=c()
se=c()
t=c()
set.seed(1234)
random_3=matrix(rnorm(reps*obs),reps,obs)
for(i in 1:reps){
  for(j in 1:obs){
    hgt[j]=random_3[i,j]*sd+mu
  }
  m[i]=mean(hgt)
  se[i]=sd(hgt)/sqrt(obs)  
  t[i]=(mu-m[i])/se[i]
}
hist(t,breaks = seq(-7,7,0.2),freq = F)
xfit=seq(-7,7,0.2)
yfit_1<-dnorm(xfit,mean=mean(t),sd=sd(t)) 
lines(xfit, yfit_1, col="blue", lwd=2) 
yfit_2<-dt(xfit,obs-1) 
lines(xfit, yfit_2, col="red", lwd=2) 
legend(x = "topright", legend=c("normal distribution", "t distribution"),
       col=c("blue", "red"),lty=c(1,1),cex=1)
})
```

For more details about Illustration of central limit theorem using Monte-Carlo simulation, see the paper of Prof.kaiji motegi 計量経済学におけるモンテカルロ実験の基本:  
<http://www2.kobe-u.ac.jp/~motegi/gakushu_motegi_v7.pdf>

## 1.2 Model estimation by OLS

### Single index model
Next, we will try to fit a univariate linear regression model for return series. 
$$
rd_t=\alpha+\beta rt_t+u_t
$$
Here are estimation results.
```{r, echo = TRUE}
eq01=lm(rd~rt)
eq01_sum=summary(eq01)
kable(eq01_sum$coefficients)
```

### t-test for parameter in eq01
Then, we can conduct t-test for prarameter in eq01 by running:
```{r echo=TRUE}
b=eq01$coefficients[2]
se=eq01_sum$coefficients[2,2]
df=eq01$df.residual
t=(b-0)/se
p=pt(t,(df-1))*2
c('t-value'=t,'p-value'=p)
```

### Coefficient of determination
$$
\begin{equation}\nonumber
    Y_i=\alpha+\beta X_i+u_i
    \end{equation}
$$
It has:
$$
\begin{equation}\nonumber
    \hat{Y}_i=\hat{\alpha}+\hat{\beta}X_i, \overline{Y}=\hat{\alpha}+\hat{\beta}\overline{X}
    \end{equation}
$$
Coefficient of determination：
$$
\begin{equation}\nonumber
R^2=\frac{\sum (\hat{Y}_i-\overline{Y})^2}{\sum (Y_i-\overline{Y})^2}
    \end{equation}
$$
Adjusted coefficient of determination：
$$
\begin{equation}\nonumber
adj R^2=1-(1-R^2)\frac{n-1}{n-k}
    \end{equation}
$$

We can check $R^2$ and $adj R^2$ by using `summary()` function easily:
```{r,echo=TRUE}
summary(eq01)
```

### Log-likelihood
The likelihood function (often simply called the likelihood) measures the goodness of fit of a statistical model to a sample of data for given values of the unknown parameters.  
```{r echo=TRUE}
set.seed(111)
random_4=matrix(rnorm(200),100,2)
x1=random_4[,1]
x2=random_4[,2]+2
den1=dnorm(x1)
den2=dnorm(x2)
logl1=sum(log(den1))
logl2=sum(log(den2))
logl1
logl2
```

Here are histogram of two series:
```{r echo=TRUE}
par(mfrow=c(1,2))
hist(x1,breaks = seq(-6,6,0.2),freq = T)
hist(x2,breaks = seq(-6,6,0.2),freq = T)
```

We can use R function `LogLik()` to obtain Log-likelihood of eq01 or calculate by ourselves.
```{r echo=TRUE}
robs=length(eq01$fitted.values)
k=2
e2=(eq01$residuals)^2
#likelihood
logl=-(robs/2)*(1+log(2*pi)+log(sum(e2)/robs))
logLik(eq01)
logl
```

### Information criterion 
### Akaike information criterion (AIC)
- AIC in Eviews:
$$
\begin{equation}\nonumber
    AIC=-2(l/T)+2(k/T)
    \end{equation}
$$    
- AIC in R:
$$
\begin{equation}\nonumber
    AIC=-2l+2k
    \end{equation}
$$    
We can use R function `AIC()` to obtain AIC of eq01 or calculate by ourselves.
As you see, the results are little different. In fact, k should be number of variables, that is 2. However, `AIC()` used `k=3`, is it a bug?
```{r echo=TRUE}
aic=-2*logl+2*k 
aic_2=AIC(eq01) 
aic_3=-2*logl/robs+2*k/robs
c(aic,aic_2,aic_3)
```

###  Bayesian information criterion (BIC)
- BIC in Eviews:
$$
\begin{equation}\nonumber
    BIC=-2(l/T)+klog(T)/T
    \end{equation}
$$    
- BIC in R:
$$
\begin{equation}\nonumber
    BIC=-2l+klog(T)
    \end{equation}
$$    
```{r echo=TRUE}
bic=-2*logl+log(robs)*k
bic_2=BIC(eq01)
bic_3=-2*logl/robs+log(robs)*k/robs
c(bic,bic_2,bic_3)
```

###  Hannan–Quinn information criterion (HQC)
- HQC in Eviews:
$$
\begin{equation}\nonumber
    HQC=-2(l/T)+2klog(log(T))/T
    \end{equation}
$$
```{r echo=TRUE}
hqc=-2*logl/robs+2*k*log(log(robs))/robs
hqc
```


### ACF and PACF
- Autocorrelation Function (ACF):
$$
\begin{equation}\nonumber
    \rho_k=\frac{\sum_{t=k+1}^{T}(Y_t-\overline{Y})(Y_{t-k}-\overline{Y})}{\sum_{t=1}^{T}(Y_t-\overline{Y})^2}
    \end{equation}
$$    
- Partial Autocorrelation Function (PACF):  
The partial autocorrelation function is a measure of the correlation between observations of a time series that are separated by k time units ($y_t$ and $y_{t–k}$), after adjusting for the presence of all the other terms of shorter lag ($y_{t–1}, y_{t–2}, ..., y_{t–k–1}$).

#### ACF and PACF of DCAM stock prices
```{r echo=TRUE}
par(mfrow=c(1,2))
acf(dcam,main='ACF plot of Stock Price of DCAM')
pacf(dcam,main='PACF plot of Stock Price of DCAM')
```

#### Try
Please check the ACF and PACF plots of `x`:
```{r g, exercise=TRUE, exercise.eval=TRUE}

```

### Ljung-Box test
The Ljung–Box test is a type of statistical test of whether any of a group of autocorrelations of a time series are different from zero.
$$
\begin{equation}\nonumber
    H_0:\rho_1=\rho_2=\dots=\rho_m=0\\
    H_1:\exists k\in [1,m], \quad s.t. \quad \rho_k \neq 0
    \end{equation}
$$
Q-statistic:
$$
\begin{equation}\nonumber
    Q(m)=T(T+2)\sum_{k=1}^{m}\frac{{\hat{\rho}_k}^2}{T-k} \sim \chi^2(m)
    \end{equation}
$$
We can conduct Ljung–Box test in R by using `Box.test()` function:
```{r echo=TRUE}
for( i in 1:13){
LB_test=Box.test(dcam_ts, lag = i, type ="Ljung-Box")
print(c(i,round(unname(LB_test$statistic),2),LB_test$p.value))
}
```

#### Try
Please check Ljung–Box test results of `x_ts` to lag 10:
```{r h, exercise=TRUE, exercise.eval=TRUE}

```

## 1.3 Estimation of AR model
### Fitting AR model by ordinary least squares
Considering a AR(1) model for DCAM stock prices:
$$
\begin{equation}\nonumber
    dcam_t=\beta_0+\beta_1dcam_{t-1}+u_t
    \end{equation}
$$  

The function `ar.ols()` allows us to fit an autoregressive time series model to the data by ordinary least squares, by default selecting the complexity by AIC in R.
```{r, echo=TRUE}
ar.ols(dcam, 
       order.max = 1, 
       demean = F, 
       intercept = T)
```       

It can be rewritten as:
$$
\begin{equation}\nonumber
    dcam_t=\alpha +u_t\\
     u_t=\rho_1u_{t-1}+\epsilon_t
    \end{equation}
$$  

### Fitting AR(1) model using AR term:
```{r, warning=FALSE, echo=TRUE}
#estimated by using conditional-sum-of-squares
AR1_1<- arima(dcam_ts, order = c(1,0,0),method='CSS')
AR1_1
```

```{r, echo=TRUE}
#estimated by using maximum likelihood
AR1_2 <- arima(dcam_ts, order = c(1,0,0),method='ML')
AR1_2
```

#### Try
Do you remember how `x_ts` was defined?  
That is:  
$$
\begin{equation}\nonumber
    x_t=0.3x_{t-1}+0.6x_{t-2}+u_{t}, u_{t} \sim N(0,16)
    \end{equation}
$$ 
please fit appropriate AR model for `x_ts` using three methods mentioned above.
```{r i, exercise=TRUE, exercise.eval=TRUE}

```


### Quiz
```{r quiz}
quiz(
  question("Which package allows us to get canlendar of a specific country?",
    answer("bizdays"),
    answer("RQuantLib", correct = TRUE),
    answer("zoo"),
    answer("Rcanlendar")
  ),
  question("What is the statistic of Ljung-Box test?",
    answer('L'),
    answer("T"),
    answer("F"),
    answer("Q", correct = TRUE)
  )
  ,
  question("MA(2) is also called?",
    answer("ARIMA(2,0,0)"),
    answer("ARIMA(0,1,2)"),
    answer("ARIMA(0,2,0)"),
    answer("ARIMA(0,0,2)",correct = TRUE)
  )
)
```

#### References

森平爽一郎, 高英模, 2019, 『Eviewで学ぶ応用ファイナンス』, 日本評論社  
羽森茂之, 2018, 『ベーシック計量経済学 第２版』, 中央経済社  
茂木快治, 2018, 計量経済学におけるモンテカルロ実験の基本, 経済学・経営学学習のために, 2018, 19-28    
沖本竜義, 2019, 『経済.ファイナンスデータの計量時系列分析』, 朝倉書店


## 1.4 Introduction of Kalman filter
If you want know about Kalman filter in advance, the youtube vedio series of matlab that introduce Kalman filter may be helpful.

![](https://www.youtube.com/watch?v=mwn8xhgNpFY&t=1s){width="70%"}
